---
title: "Fifty years of Vancouver street network"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fifty years of Vancouver street network}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 5, dpi = 96, out.width = "100%",
                      fig.path = "canstreet-vancouver-")
```

The Calgary vignette explains a build; this one reads one. It is the eleven
census years from 1971 to 2021, in four formats --- mainframe flat files no GIS
driver opens, Street Network File coverages, a MapInfo interchange pair, and
Road Network File shapefiles --- which the package makes irrelevant to the
caller. Vancouver is the region to read it over because its Area Master Files
reach well beyond the city from the start, twelve municipal files of them in
1971, so half a century of one metropolitan area is in view from the first
year.


```{r library, message = FALSE}
library(canstreet)
library(dplyr)
library(ggplot2)
```

## A region to work in

As in the Calgary vignette, one fixed polygon is stamped across every year
rather than filtering on a region attribute, which the pre-2011 files do not
carry anyway:

```{r region}
cma <- cancensus::get_statcan_geographies("2021", level = "CMA", type="digital") |>
  filter(CMANAME == "Vancouver")

vancouver_crs <- "+proj=lcc +lat_1=49.0000658247404 +lat_2=49.5742753405107 +lat_0=49.2871705826255 +lon_0=-123.073508713002 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"

cma |> sf::st_drop_geometry() |> select(CMAUID, CMANAME)
```

## Overview

Before 2001's national reach the files hold only the larger urban areas, and
which areas counted changed from vintage to vintage. Eleven reads, varying only
the year:

```{r overview, fig.height = 8.2, fig.alt = "The Vancouver CMA road network as each of the eleven vintages holds it, 1971 to 2021, showing coverage expanding from a few urban areas to the whole region"}
vints <- c(1971, 1976, 1981, 1986, 1991, 1996, 2001, 2006, 2011, 2016, 2021)

overview <- vints |>
  lapply(function(y) {
    get_road_network(y, within = cma, roads_only = TRUE, quiet = TRUE) |>
      select(vintage, geom) |>
      collect_road_network()
  }) |>
  (\(x) do.call(rbind, x))()

ggplot(overview |> sf::st_transform(vancouver_crs)) +
  geom_sf(data = cma |> sf::st_transform(vancouver_crs),
          fill = "grey94", colour = NA) +
  geom_sf(linewidth = 0.05) +
  facet_wrap(~vintage, ncol = 3) +
  coord_sf(datum = NA) +
  labs(title = "Vancouver road network over time",
       subtitle = "Each panel is that vintage's own file, over the 2021 CMA",
       caption = "Source: Statistics Canada Area Master, Street Network and Road Network Files")
```

Most of what fills in before 2001 is coverage, not construction --- the eastern
end of the region arrives all at once in 1986. Keep that in mind for
`first_year` below.



## Building

The Area Master Files join the series like any other vintage:

```{r build, message = FALSE}

vancouver <- build_temporal_network(
  "vancouver", vints, within = cma,
  region_note = "2021 CMA boundary, Vancouver (933)") 

vancouver
```

## Read `first_year` as "first mapped", not "first built"

`first_year` is the earliest vintage a segment was matched in --- a statement
about the files, which do not all cover the same ground:

```{r coverage}
list_road_network_vintages() |>
  filter(vintage %in% vints) |>
  select(vintage, product, coverage, n_segments)
```

A road in Langley that first appears in 1986 may well have been there in 1971,
simply outside what anyone had digitized. Growth and coverage are confounded,
and matching cannot separate them. The kilometres present in each year stack
both effects:

```{r by-year}
tibble::tibble(year = vints, km = round(vapply(vints, function(y) {
  vancouver |>
    filter(list_contains(years, y)) |>
    summarize(km = sum(len_m, na.rm = TRUE) / 1000) |>
    pull(km)
}, numeric(1)), 1))
```

## Where the confounding goes away

Inside a municipality already fully mapped in 1971 the coverage problem does not
arise. Subset it with a polygon, not an attribute: `csduid_l` comes from the
vintage a segment's geometry was cut from, and the files before 2011 carry no
region columns, so an attribute filter would drop exactly the oldest segments.

```{r core}
van_city <- cancensus::get_statcan_geographies("2021", level = "CSD", type = "digital") |>
  filter(CSDUID == "5915022")

city <- get_temporal_network("vancouver", within = van_city)

by_first <- function(x) {
  x |>
    group_by(first_year) |>
    summarize(km = sum(len_m, na.rm = TRUE) / 1000) |>
    collect() |>
    arrange(first_year) |>
    mutate(share = round(km / sum(km), 3), km = round(km, 1))
}

by_first(vancouver) |>
  rename(km_cma = km, share_cma = share) |>
  left_join(by_first(city) |> rename(km_city = km, share_city = share),
            by = "first_year")
```

In the city 93% of the length was already mapped in 1971 --- the grid was
substantially complete, and what came later is infill. CMA-wide it is 56%, with
1986 alone adding 1,850 km: that is the Area Master File taking in Langley,
Maple Ridge and Pitt Meadows for the first time, not road being built.


## The map

The concentric structure is real growth; the hard edges are where a file's
coverage stopped. Only segments still present in 2021 are drawn.

```{r first-year-map, fig.alt = "Vancouver CMA road network coloured by the first vintage each segment appears in, 1971 to 2021"}
plot_first_year <- function(x, title, linewidth = 0.1) {
  colours <- c("grey40",viridis::viridis(length(vints)-1, option = "turbo", direction = -1,
                           begin = 0.12, end = 0.92))
  x |>
    filter(list_contains(years, 2021L)) |>
    collect_road_network() |>
    sf::st_transform(vancouver_crs) |>
    ggplot(aes(colour = factor(first_year))) +
    geom_sf(linewidth = linewidth) +
    scale_colour_manual(values = colours) +
    labs(title = title,
         subtitle = "Colour is the first year the segment appears in the data",
         colour = "First year\nroad appears\nin data",
         caption = "StatCan Area Master, Street Network and Road Network Files") +
    guides(colour = guide_legend(override.aes = list(linewidth = 1))) +
    coord_sf(datum = NA)
}

plot_first_year(vancouver, "Vancouver CMA road network by vintage")
```

Zooming to the City of Vancouver takes the coverage edges out of the frame, so
what remains is close to a map of when the city's streets were built.

```{r core-map, fig.alt = "City of Vancouver streets coloured by the first vintage each segment appears in"}
plot_first_year(city, "City of Vancouver streets by vintage",
                linewidth = 0.25)
```

## Matching across fifty years

The Area Master Files might be expected to be the weak link --- hand-digitized
map sheets, NAD27 UTM coordinates, a record layout from the mainframe era. The
calibration says otherwise:

```{r calibration}
temporal_network_calibration("vancouver")
```

`recall_p50` is the positional disagreement against 2021, measured on roads the
two files name alike. The four Area Master Files come in between 10.8 m and
11.4 m --- *better* than 1991 and 1996 at 15.6 m and 15.7 m, and on a par with
2001 and 2006. What improves after 2006 is vertex density rather than
registration; 2011 and 2016 share 2021's base geometry, which is why they
calibrate to the 10 m floor.

That is Vancouver's result and not a property of the product. Over Calgary the
same four years disagree with 2021 by 17 m to 20 m and sit systematically to one
side of it, which the build there warns about --- see
`vignette("canstreet-temporal")`.

The crosswalk records how each year's segments were matched:

```{r crosswalk}
get_temporal_network_sources("vancouver") |>
  count(vintage, match_kind) |>
  arrange(vintage, desc(n)) |>
  collect()
```

`name_rescue` is the long-range same-name match, for arterials an older file
draws as a coarse chord the modern geometry departs from. It is recorded rather
than folded in, so a stricter reading is a filter away.

## The raw Area Master File

No GIS driver opens the Area Master File --- fixed-width mainframe records, one
flat file per municipality, the coordinates written out as digits --- so the
package parses it. `read_amf()` exposes it directly, as segments or as the node
records the file stores:

```{r read-amf}
amf <- canstreet_download(1971)
files <- utils::unzip(amf$path, exdir = tempdir())
nodes <- read_amf(grep("VANCOUVE", files, value = TRUE), nodes = TRUE)

nodes |>
  select(sheet_name, feature, seq_no, name, type, class, x, y, zone) |>
  head(5)
```

A node carries the feature's name and type, a class where it is not an ordinary
street, the civic address range on each side, and a NAD27 UTM coordinate in its
map sheet's zone. A segment is the block face between two consecutive nodes ---
median length 109 m in this file, finer than the later files use.

## Attribution

Source data are © Statistics Canada, distributed under the
[Statistics Canada Open Licence](https://www.statcan.gc.ca/en/reference/licence).
Cite the product and reference year in anything you publish from it.
