---
title: "Forty-five years of Vancouver street network"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forty-five years of Vancouver street network}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 5, dpi = 96, out.width = "100%",
                      fig.path = "canstreet-vancouver-")
```

The Calgary vignette builds a temporal network from the census years 1996
onward. Vancouver reaches two decades further back, because the two Area Master
Files Statistics Canada deposited for 1976 and 1981 cover British Columbia and
nothing else. That makes nine vintages over forty-five years --- and they are
not variations on a single format. 1976 and 1981 are mainframe flat files that
no GIS driver opens, 1991 and 1996 are Street Network File shapefiles, 2001 is
a MapInfo interchange pair, and 2006 onward are Road Network File shapefiles. The
package's job is to make that irrelevant to the caller.

This vignette is precomputed --- it downloads several hundred megabytes and the
build takes a few minutes --- but the code is exactly what you would run.

```{r library}
library(canstreet)
library(dplyr)
library(ggplot2)
```

## A region to work in

As in the Calgary vignette, one fixed polygon is stamped across every year
rather than filtering on a region attribute, which the pre-2011 files do not
carry anyway:

```{r region}
cma <- cancensus::get_statcan_geographies("2021", level = "CMA") |>
  filter(CMANAME == "Vancouver")

cma |> sf::st_drop_geometry() |> select(CMAUID, CMANAME)
```

## Building

The two Area Master Files join the series like any other vintage. Nothing in
the call says they arrive in a different format:

```{r build, message = FALSE}
vints <- c(1976, 1981, 1991, 1996, 2001, 2006, 2011, 2016, 2021)

vancouver <- build_temporal_network(
  "vancouver", vints, within = cma,
  region_note = "2021 CMA boundary, Vancouver (933)")

vancouver
```

## Read `first_year` as "first mapped", not "first built"

This is the caution the whole vignette turns on, and it gets sharper the
further back the series reaches.

`first_year` is the earliest vintage in which a segment was matched. That is a
statement about the *files*, and the files do not all cover the same ground:

```{r coverage}
list_road_network_vintages() |>
  filter(vintage %in% vints) |>
  select(vintage, product, coverage, n_segments)
```

1976 and 1981 map a handful of British Columbia urban areas; 1991 and 1996 map
urban areas nationally; 2001 is the first file covering the whole country. So a
road in Langley or Maple Ridge that first appears in 2001 may well have been
there in 1976 --- it was simply outside what anyone had digitized. Growth and
coverage are confounded, and no amount of matching separates them.

The kilometres present in each year show the two effects stacked:

```{r by-year}
tibble::tibble(year = vints, km = round(vapply(vints, function(y) {
  vancouver |>
    filter(list_contains(years, y)) |>
    summarize(km = sum(len_m, na.rm = TRUE) / 1000) |>
    pull(km)
}, numeric(1)), 1))
```

## Where the confounding goes away

Inside a municipality that was already fully mapped in 1976, the coverage
problem does not arise and `first_year` recovers its plain meaning. The City of
Vancouver is such a place.

Subsetting it is the same operation as building the region in the first place
--- a polygon, not an attribute. That is not fussiness: `csduid_l` on a segment
comes from the vintage its geometry was cut from, and the 1976 through 2006
files carry no region columns at all, so filtering on the attribute would
silently drop exactly the oldest segments this vignette is about.

```{r core}
van_city <- cancensus::get_statcan_geographies("2021", level = "CSD") |>
  filter(CSDUID == "5915022")

city <- get_temporal_network("vancouver", within = van_city)

by_first <- function(x) {
  x |>
    group_by(first_year) |>
    summarize(km = sum(len_m, na.rm = TRUE) / 1000) |>
    collect() |>
    arrange(first_year) |>
    mutate(share = round(km / sum(km), 3), km = round(km, 1))
}

by_first(vancouver) |>
  rename(km_cma = km, share_cma = share) |>
  left_join(by_first(city) |> rename(km_city = km, share_city = share),
            by = "first_year")
```

The two columns tell different stories. In the city 91% of the network's length
was already mapped in 1976, which is the honest reading: the street grid was
substantially complete by then, and what came later is infill --- a few
kilometres a decade. Across the CMA it is 57%, and the 1991 column alone adds
2,435 km. Vancouver did not gain 2,435 km of road in 1991. That is the year the
Street Network File extended past the 1981 urban envelope, and it is the
coverage effect in a single number.

## The map

Colouring each segment by the year it first appears makes both readings visible
at once. The concentric structure is real urban growth; the hard edges are
where a file's coverage stopped. Only segments still present in 2021 are drawn.

```{r first-year-map, fig.alt = "Vancouver CMA road network coloured by the first vintage each segment appears in, 1976 to 2021"}
plot_first_year <- function(x, title, linewidth = 0.1) {
  x |>
    filter(list_contains(years, 2021L)) |>
    collect_road_network() %>%
    sf::st_transform(mountainmathHelpers::lambert_conformal_conic_at(.)) |>
    ggplot(aes(colour = factor(first_year))) +
    geom_sf(linewidth = linewidth) +
    scale_colour_viridis_d(option = "turbo", direction = -1,
                           begin = 0.12, end = 0.92) +
    labs(title = title,
         subtitle = "Colour is the first year the segment appears in the data",
         colour = "First year\nroad appears\nin data",
         caption = "StatCan Area Master, Street Network and Road Network Files") +
    guides(colour = guide_legend(override.aes = list(linewidth = 1))) +
    coord_sf(datum = NA)
}

plot_first_year(vancouver, "Vancouver CMA road network by vintage")
```

Zooming to the City of Vancouver takes the coverage edges out of the frame
entirely, so what remains is close to a map of when the city's streets were
built --- or, at the margins, of when a lane or a cul-de-sac was first thought
worth digitizing.

```{r core-map, fig.alt = "City of Vancouver streets coloured by the first vintage each segment appears in"}
plot_first_year(city, "City of Vancouver streets by vintage",
                linewidth = 0.25)
```

## Matching across forty-five years

The expectation going in was that the Area Master Files would be the weak link:
hand-digitized from map sheets, stored as NAD27 UTM coordinates, packed
two decimal digits to the byte. The calibration says otherwise.

```{r calibration}
temporal_network_calibration("vancouver")
```

`recall_p50` is the everyday positional disagreement against 2021, measured on
roads the two files agree on the name of. 1976 and 1981 come in at 11.5 m and
11.0 m --- *better* than the 1991 and 1996 shapefiles at 15.7 m, and on a par
with 2001 and 2006. The 1970s map sheets were drawn carefully; what improves
after 2006 is not so much accuracy as the density of vertices, and 2011 and
2016 share their base geometry with 2021 outright, which is why they calibrate
to the 10 m floor.

The crosswalk records how each year's segments were matched:

```{r crosswalk}
get_temporal_network_sources("vancouver") |>
  count(vintage, match_kind) |>
  arrange(vintage, desc(n)) |>
  collect()
```

`name_rescue` is the same-name match at long range that exists because older
files draw a long arterial as a coarse chord the modern geometry departs from.
It carries a real share of the two oldest vintages, and it is recorded rather
than folded in, so a stricter reading is a filter away.

## The raw Area Master File

The two oldest vintages reach the database through a parser written for this
package, because no GIS driver opens the format: fixed-width mainframe records,
113 bytes in 1976 and 95 in 1981, the second still in EBCDIC with
packed-decimal coordinates. `read_amf()` exposes it directly, as segments or as
the node records the file actually stores:

```{r read-amf}
amf <- canstreet_download(1976)
nodes <- read_amf(amf$path[amf$filename == "vancouver.data"], nodes = TRUE)

nodes |>
  select(sheet_name, feature, seq_no, name, type, class, x, y, zone) |>
  head(5)
```

A node carries its feature's name and type, a class where the feature is
something other than an ordinary street, the civic address range on each side,
and a NAD27 UTM coordinate in the zone its map sheet belongs to. A segment is
the block face between two consecutive nodes of a chain --- a finer unit than
the later files use, median length about 100 m, which is why 1976 contributes
as many segments as it does over so small an area.

## Attribution

Source data are © Statistics Canada, distributed under the
[Statistics Canada Open Licence](https://www.statcan.gc.ca/en/reference/licence).
Cite the product and reference year in anything you publish from it.
