---
title: "Getting started with canstreet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with canstreet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Figures are written beside the .Rmd rather than into a `figure/` directory:
# R CMD check reports a directory by that name in vignettes/ as knitr detritus,
# but a precomputed vignette has to ship its figures.
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96, out.width = "100%",
                      fig.path = "canstreet-")
```

This vignette is precomputed: it downloads several hundred megabytes and takes
a few minutes, which is more than a package build should do. The code is
exactly what you would run yourself.

```{r library}
library(canstreet)
library(dplyr)
```

## Set a cache path

Road network files are large — a national vintage is 200–350 MB compressed and
about 2.2 million segments — so `canstreet` caches both the downloaded archives
and the harmonized database. Set a directory once and it applies to every
session:

```{r cache, eval = FALSE}
set_canstreet_cache_path("~/data/canstreet", install = TRUE)
```

Without one, everything goes to `tempdir()` and is downloaded again next time.
`show_canstreet_cache_path()` reports what is in effect and where the setting
came from.

## What is available

```{r vintages}
list_road_network_vintages() |>
  select(vintage, product, coverage, source_crs, imported)
```

The 1991 and 1996 Street Network Files cover urban areas only; everything from
2001 on is national.

## Getting a vintage

The first call downloads and imports; after that it is instant.

```{r get}
roads <- get_road_network(2021)
roads
```

The result is a lazy table over DuckDB, so this aggregation over 2.2 million
segments runs inside the database and only the answer comes back to R:

```{r by-province}
roads |>
  filter(!is.na(pruid_l)) |>
  group_by(pruid_l) |>
  summarize(segments = n(), km = round(sum(len_m, na.rm = TRUE) / 1000)) |>
  arrange(desc(km)) |>
  collect()
```

## Spatial filters

Pass any `sf`, `sfc` or `bbox` object as `within`. Over a single vintage the
filter uses that vintage's R-tree index, so it stays fast on a national file.
`collect_road_network()` then materializes the result as `sf`.

```{r vancouver}
bbox <- sf::st_bbox(c(xmin = -123.15, ymin = 49.26, xmax = -123.09,
                      ymax = 49.29), crs = 4326)

downtown <- get_road_network(2021, within = bbox) |>
  collect_road_network()

downtown |> select(name, type, af_l, at_l, csdname_l)
```

```{r plot, fig.alt = "Street segments in downtown Vancouver, 2021"}
plot(sf::st_geometry(downtown), col = "grey30", lwd = 0.6)
```

Geometry is stored in EPSG:3347 (NAD83 / Statistics Canada Lambert), which is
why `len_m` is in metres for every vintage. Ask
`collect_road_network()` for another CRS when you need one:

```{r crs}
sf::st_crs(collect_road_network(get_road_network(2021, within = bbox),
                                crs = 4326))$epsg
```

## Comparing vintages

Several years at once are stacked into one table, which is what makes change
over time a single query:

```{r across-years}
get_road_network(c(1996, 2006, 2021)) |>
  group_by(vintage) |>
  summarize(segments = n(), km = round(sum(len_m, na.rm = TRUE) / 1000)) |>
  collect()
```

Read those totals with the coverage difference in mind: 1996 is urban areas
only. The 2006–2021 drop is in the source data rather than in the import — the
2006 file carries far more very long arcs, which later releases split,
reclassified or dropped:

```{r long-arcs}
get_road_network(c(2006, 2021)) |>
  filter(len_m > 5000) |>
  group_by(vintage) |>
  summarize(long_arcs = n(), km = round(sum(len_m, na.rm = TRUE) / 1000)) |>
  collect()
```

It is also why matching segments across years is its own piece of work rather
than a join on an identifier: identifiers are not documented as stable, and
Statistics Canada states arc positions are not necessarily consistent between
versions. `build_temporal_network()` does that matching and writes a single
table in which each segment carries the years it is present in --- see
`vignette("canstreet-temporal")`.

## The harmonized schema

Every vintage is mapped onto the same columns. A column that a vintage's source
files do not carry is `NA` throughout that vintage rather than absent, so the
union stays legal.

```{r schema}
canstreet_schema()
```

Vintages spell things very differently — `arc_id` in 1991/1996, `RB_UID`
in 2001–2010, `NGD_UID` from 2011 and `NGDUID` in 2016/2017 — and the importer
resolves each by name, case-insensitively, rather than assuming an era.

## Exporting

`export_road_network()` writes GeoParquet with the CRS in its metadata, for use
from Python, QGIS or DuckDB:

```{r export, eval = FALSE}
get_road_network(2021, within = bbox) |>
  export_road_network("vancouver_2021.parquet")
```

## Managing the cache

```{r cache-listing}
list_canstreet_cache()
```

`remove_canstreet_cache(vintage)` drops a vintage's table and keeps its
archives, so it can be re-imported without downloading again; pass
`keep_raw = FALSE` to delete those too.

## Attribution

Source data are © Statistics Canada, distributed under the
[Statistics Canada Open Licence](https://www.statcan.gc.ca/en/reference/licence).
Cite the product and reference year in anything you publish from it.
