---
title: "Forty-five years of Vancouver street network"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forty-five years of Vancouver street network}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 5, dpi = 96, out.width = "100%",
                      fig.path = "canstreet-vancouver-")
```

Vancouver reaches two decades further back than the Calgary vignette, because
the two Area Master Files Statistics Canada deposited for 1976 and 1981 cover
British Columbia. That is nine vintages over forty-five years in four formats
--- mainframe flat files no GIS driver opens, Street Network File coverages, a
MapInfo interchange pair, and Road Network File shapefiles --- which the package
makes irrelevant to the caller.


```{r library}
library(canstreet)
library(dplyr)
library(ggplot2)
```

## A region to work in

As in the Calgary vignette, one fixed polygon is stamped across every year
rather than filtering on a region attribute, which the pre-2011 files do not
carry anyway:

```{r region}
cma <- cancensus::get_statcan_geographies("2021", level = "CMA", type="digital") |>
  filter(CMANAME == "Vancouver")

vancouver_crs <- "+proj=lcc +lat_1=49.0000658247404 +lat_2=49.5742753405107 +lat_0=49.2871705826255 +lon_0=-123.073508713002 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"

cma |> sf::st_drop_geometry() |> select(CMAUID, CMANAME)
```

## Overview

Before 2001's national reach the files hold only the larger urban areas, and
which areas counted changed from vintage to vintage. Nine reads of nine
products, varying only the year:

```{r overview, fig.height = 8, fig.alt = "The Vancouver CMA road network as each of the nine vintages holds it, 1976 to 2021, showing coverage expanding from a few urban areas to the whole region"}
vints <- c(1976, 1981, 1991, 1996, 2001, 2006, 2011, 2016, 2021)

overview <- vints |>
  lapply(function(y) {
    get_road_network(y, within = cma, roads_only = TRUE, quiet = TRUE) |>
      select(vintage, geom) |>
      collect_road_network()
  }) |>
  (\(x) do.call(rbind, x))()

ggplot(overview |> sf::st_transform(vancouver_crs)) +
  geom_sf(data = cma |> sf::st_transform(vancouver_crs),
          fill = "grey94", colour = NA) +
  geom_sf(linewidth = 0.05) +
  facet_wrap(~vintage, ncol = 3) +
  coord_sf(datum = NA) +
  labs(title = "Vancouver road network over time",
       subtitle = "Each panel is that vintage's own file, over the 2021 CMA",
       caption = "Source: Statistics Canada Area Master, Street Network and Road Network Files")
```

Most of what fills in before 2001 is coverage, not construction --- keep that in
mind for `first_year` below.



## Building

The Area Master Files join the series like any other vintage:

```{r build, message = FALSE}

vancouver <- build_temporal_network(
  "vancouver", vints, within = cma,
  region_note = "2021 CMA boundary, Vancouver (933)") 

vancouver
```

## Read `first_year` as "first mapped", not "first built"

`first_year` is the earliest vintage a segment was matched in --- a statement
about the files, which do not all cover the same ground:

```{r coverage}
list_road_network_vintages() |>
  filter(vintage %in% vints) |>
  select(vintage, product, coverage, n_segments)
```

A road in Langley that first appears in 2001 may well have been there in 1976,
simply outside what anyone had digitized. Growth and coverage are confounded,
and matching cannot separate them. The kilometres present in each year stack
both effects:

```{r by-year}
tibble::tibble(year = vints, km = round(vapply(vints, function(y) {
  vancouver |>
    filter(list_contains(years, y)) |>
    summarize(km = sum(len_m, na.rm = TRUE) / 1000) |>
    pull(km)
}, numeric(1)), 1))
```

## Where the confounding goes away

Inside a municipality already fully mapped in 1976 the coverage problem does not
arise. Subset it with a polygon, not an attribute: `csduid_l` comes from the
vintage a segment's geometry was cut from, and the 1976--2006 files carry no
region columns, so an attribute filter would drop exactly the oldest segments.

```{r core}
van_city <- cancensus::get_statcan_geographies("2021", level = "CSD", type = "digital") |>
  filter(CSDUID == "5915022")

city <- get_temporal_network("vancouver", within = van_city)

by_first <- function(x) {
  x |>
    group_by(first_year) |>
    summarize(km = sum(len_m, na.rm = TRUE) / 1000) |>
    collect() |>
    arrange(first_year) |>
    mutate(share = round(km / sum(km), 3), km = round(km, 1))
}

by_first(vancouver) |>
  rename(km_cma = km, share_cma = share) |>
  left_join(by_first(city) |> rename(km_city = km, share_city = share),
            by = "first_year")
```

In the city 93% of the length was already mapped in 1976 --- the grid was
substantially complete, and what came later is infill. CMA-wide it is 57%, with
1991 alone adding 2,421 km: that is the Street Network File reaching past the
1981 urban envelope, not road being built.


## The map

The concentric structure is real growth; the hard edges are where a file's
coverage stopped. Only segments still present in 2021 are drawn.

```{r first-year-map, fig.alt = "Vancouver CMA road network coloured by the first vintage each segment appears in, 1976 to 2021"}
plot_first_year <- function(x, title, linewidth = 0.1) {
  x |>
    filter(list_contains(years, 2021L)) |>
    collect_road_network() |>
    sf::st_transform(vancouver_crs) |>
    ggplot(aes(colour = factor(first_year))) +
    geom_sf(linewidth = linewidth) +
    scale_colour_viridis_d(option = "turbo", direction = -1,
                           begin = 0.12, end = 0.92) +
    labs(title = title,
         subtitle = "Colour is the first year the segment appears in the data",
         colour = "First year\nroad appears\nin data",
         caption = "StatCan Area Master, Street Network and Road Network Files") +
    guides(colour = guide_legend(override.aes = list(linewidth = 1))) +
    coord_sf(datum = NA)
}

plot_first_year(vancouver, "Vancouver CMA road network by vintage")
```

Zooming to the City of Vancouver takes the coverage edges out of the frame, so
what remains is close to a map of when the city's streets were built.

```{r core-map, fig.alt = "City of Vancouver streets coloured by the first vintage each segment appears in"}
plot_first_year(city, "City of Vancouver streets by vintage",
                linewidth = 0.25)
```

## Matching across forty-five years

The Area Master Files might be expected to be the weak link --- hand-digitized
map sheets, NAD27 UTM coordinates, packed two decimal digits to the byte. The
calibration says otherwise:

```{r calibration}
temporal_network_calibration("vancouver")
```

`recall_p50` is the positional disagreement against 2021, measured on roads the
two files name alike. 1976 and 1981 come in at 11.5 m and 11.0 m --- *better*
than 1991 and 1996 at 15.7 m, and on a par with 2001 and 2006. What improves
after 2006 is vertex density rather than registration; 2011 and 2016 share
2021's base geometry, which is why they calibrate to the 10 m floor.

The crosswalk records how each year's segments were matched:

```{r crosswalk}
get_temporal_network_sources("vancouver") |>
  count(vintage, match_kind) |>
  arrange(vintage, desc(n)) |>
  collect()
```

`name_rescue` is the long-range same-name match, for arterials an older file
draws as a coarse chord the modern geometry departs from. It is recorded rather
than folded in, so a stricter reading is a filter away.

## The raw Area Master File

No GIS driver opens the Area Master File --- fixed-width mainframe records, 113
bytes in 1976 and 95 in 1981, the second in EBCDIC with packed-decimal
coordinates --- so the package parses it. `read_amf()` exposes it directly, as
segments or as the node records the file stores:

```{r read-amf}
amf <- canstreet_download(1976)
nodes <- read_amf(amf$path[amf$filename == "vancouver.data"], nodes = TRUE)

nodes |>
  select(sheet_name, feature, seq_no, name, type, class, x, y, zone) |>
  head(5)
```

A node carries the feature's name and type, a class where it is not an ordinary
street, the civic address range on each side, and a NAD27 UTM coordinate in its
map sheet's zone. A segment is the block face between two consecutive nodes ---
median length about 100 m, finer than the later files use.

## Attribution

Source data are © Statistics Canada, distributed under the
[Statistics Canada Open Licence](https://www.statcan.gc.ca/en/reference/licence).
Cite the product and reference year in anything you publish from it.
