---
title: "Getting started with canstreet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with canstreet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Figures are written beside the .Rmd rather than into a `figure/` directory:
# R CMD check reports a directory by that name in vignettes/ as knitr detritus,
# but a precomputed vignette has to ship its figures.
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96, out.width = "100%",
                      fig.path = "canstreet-")
```

This vignette is precomputed --- it downloads several hundred megabytes --- but
the code is what you would run.

```{r library}
library(canstreet)
library(dplyr)
```

## Set a cache path

A national vintage is 200–350 MB compressed and about 2.2 million segments, so
`canstreet` caches the archives and the harmonized database. Set a directory
once and it applies to every session:

```{r cache, eval = FALSE}
set_canstreet_cache_path("~/data/canstreet", install = TRUE)
```

Without one everything goes to `tempdir()`. `show_canstreet_cache_path()`
reports what is in effect.

## What is available

```{r vintages}
list_road_network_vintages() |>
  select(vintage, product, coverage, source_crs, imported)
```

The 1991 and 1996 Street Network Files cover urban areas only; everything from
2001 on is national.

## Getting a vintage

The first call downloads and imports; after that it is instant.

```{r get}
roads <- get_road_network(2021)
roads
```

The result is a lazy table over DuckDB, so this aggregation over 2.2 million
segments runs in the database and only the answer comes back:

```{r by-province}
roads |>
  filter(!is.na(pruid_l)) |>
  group_by(pruid_l) |>
  summarize(segments = n(), km = round(sum(len_m, na.rm = TRUE) / 1000)) |>
  arrange(desc(km)) |>
  collect()
```

## Spatial filters

Pass any `sf`, `sfc` or `bbox` object as `within`. Over a single vintage the
filter uses that vintage's R-tree index, so it stays fast on a national file.
`collect_road_network()` then materializes the result as `sf`.

```{r vancouver}
bbox <- sf::st_bbox(c(xmin = -123.15, ymin = 49.26, xmax = -123.09,
                      ymax = 49.29), crs = 4326)

downtown <- get_road_network(2021, within = bbox) |>
  collect_road_network()

downtown |> select(name, type, af_l, at_l, csdname_l)
```

```{r plot, fig.alt = "Street segments in downtown Vancouver, 2021"}
plot(sf::st_geometry(downtown), col = "grey30", lwd = 0.6)
```

Geometry is stored in EPSG:3347 (NAD83 / Statistics Canada Lambert), so `len_m`
is metres for every vintage. `collect_road_network()` reprojects on request:

```{r crs}
sf::st_crs(collect_road_network(get_road_network(2021, within = bbox),
                                crs = 4326))$epsg
```

## Comparing vintages

Several years at once stack into one table:

```{r across-years}
get_road_network(c(1996, 2006, 2021)) |>
  group_by(vintage) |>
  summarize(segments = n(), km = round(sum(len_m, na.rm = TRUE) / 1000)) |>
  collect()
```

1996 is urban areas only. The 2006–2021 drop is in the source data, not the
import: 2006 carries far more very long arcs, which later releases split,
reclassified or dropped.

```{r long-arcs}
get_road_network(c(2006, 2021)) |>
  filter(len_m > 5000) |>
  group_by(vintage) |>
  summarize(long_arcs = n(), km = round(sum(len_m, na.rm = TRUE) / 1000)) |>
  collect()
```

Identifiers are not documented as stable across vintages, so comparing years
means matching geometry. `build_temporal_network()` does that and writes one
table in which each segment carries the years it is present in --- see
`vignette("canstreet-temporal")`.

## The harmonized schema

Every vintage maps onto the same columns; one a vintage's files do not carry is
`NA` throughout it rather than absent, so the union stays legal.

```{r schema}
canstreet_schema()
```

Source spellings differ by vintage — `arc_id` in 1991/1996, `RB_UID` in
2001–2010, `NGD_UID` from 2011, `NGDUID` in 2016/2017 — and the importer
resolves each by alias, not by era.

## Exporting

`export_road_network()` writes GeoParquet with the CRS in its metadata, for use
from Python, QGIS or DuckDB:

```{r export, eval = FALSE}
get_road_network(2021, within = bbox) |>
  export_road_network("vancouver_2021.parquet")
```

## Managing the cache

```{r cache-listing}
list_canstreet_cache()
```

`remove_canstreet_cache(vintage)` drops a vintage's table and keeps its
archives, so it can be re-imported without downloading again; pass
`keep_raw = FALSE` to delete those too.

## Attribution

Source data are © Statistics Canada, distributed under the
[Statistics Canada Open Licence](https://www.statcan.gc.ca/en/reference/licence).
Cite the product and reference year in anything you publish from it.
