---
title: "Street renaming in the City of Vancouver"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Street renaming in the City of Vancouver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 5, dpi = 96, out.width = "100%",
                      fig.path = "canstreet-renames-")
```

To showcase what else can be done with the **canstreet** package we give an example of systematically detecting renaming of roads. For this we take the temporal matching of roads and compare the names of the associated segments in each year. To be useful this needs to distinguish changes in writing of the same name from actual renamings. Most of what a naive diff returns is Statistics Canada changing how names were written. This vignette peels those layers off in order, over one city, until what is left is worth reading.

```{r library, message = FALSE}
library(canstreet)
library(dplyr)
library(ggplot2)
```

## The region

We pick Vancouver for this example and re-use the temporal build at the CMA to look at road names at the city level.

```{r build, message = FALSE}
cma <- cancensus::get_statcan_geographies("2021", level = "CMA", type = "digital") |>
  filter(CMANAME == "Vancouver")

van_city <- cancensus::get_statcan_geographies("2021", level = "CSD", type = "digital") |>
  filter(CSDUID == "5915022")

vancouver_crs <- "+proj=lcc +lat_1=49.0000658247404 +lat_2=49.5742753405107 +lat_0=49.2871705826255 +lon_0=-123.073508713002 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"

vancouver <- build_temporal_network(
  "vancouver", c(1971, 1976, 1981, 1986, 1991, 1996, 2001, 2006, 2011, 2016, 2021),
  within = cma, region_note = "2021 CMA boundary, Vancouver (933)")

city <- get_temporal_network("vancouver", within = van_city) |>
  select(segment_id, first_year, len_m) |>
  collect()

sources <- get_temporal_network_sources("vancouver")
```

To subset we use the geographic extent instead of the `csduid_l` attributes, as geographic boundaries of the City of Vancouver have shifted over this timeframe. Moreover, the files before 2011 do not carry region columns.

## Street names over time

A crosswalk row is one segment in one year, and `src_name` is the respective name for the given year. As an example we select a segment that was renamed twice over this time span.

```{r one-segment}
# Looked up by what three of the files call it: a `segment_id` belongs to one
# build and does not survive a rebuild from different sources.
called <- function(year, nm) {
  sources |> filter(vintage == year, src_name == nm) |> distinct(segment_id)
}

kent_block <- called(1971, "KENT") |>
  inner_join(called(1996, "NORTH KENT"), by = "segment_id") |>
  inner_join(called(2021, "E Kent"), by = "segment_id") |>
  summarize(segment_id = min(segment_id, na.rm = TRUE)) |>
  pull(segment_id)

sources |>
  filter(segment_id == kent_block) |>
  select(vintage, src_name, match_kind, dist_m, source_file) |>
  collect() |>
  arrange(vintage)
```

That is a block of Kent Avenue on the Fraser, called `KENT` in every file from
1971 to 2011 but one, `NORTH KENT` in 1996 alone, and `E Kent` from 2016 on.

Note that `source_id` is not necessarily unique within a `vintage`: for older
vintages the source data was split into several regional files, each with its
own `source_id` numbering, so it takes `(source_file, source_id)` to uniquely identify an arc of a given
year.

And `match_kind` is `geometry` on every row before 2016. `spine` is the year the
geometry came from, `geometry+name` is a match where the two files agreed on the
name, and **`geometry` is a match where names did not match** --- the arcs sit within
the calibrated tolerance and run the same way, but are named differently. That
is the rename signal we are targeting here.

## Sequencing the names

To track renamings we order each segment's rows by year and compare each name
with the one before it. We do the same on folded names, which catch common
changes in spelling that do not correspond to an underlying name change. DuckDB
lets us do this efficiently in the database.


```{r steps}
steps <- sources |>
  mutate(src_name=coalesce(src_name,"")) |>
  mutate(nm = regexp_replace(strip_accents(toupper(src_name)),
                             '^([0-9]+)(ST|ND|RD|TH)$', '\\1')) |>
  select(segment_id, vintage, src_name, nm, dist_m) |>
  filter(segment_id %in% city$segment_id) |>
  mutate(prev_name = lag(src_name, order_by = vintage), 
         prev_nm = lag(nm, order_by = vintage),
         prev_year = lag(vintage, order_by = vintage), 
         prev_dist = lag(dist_m, order_by = vintage),
         .by=segment_id) |>
  filter(!is.na(prev_name)) |>
  collect() |>
  left_join(city |> select(segment_id, len_m), by = "segment_id")

filled <- function(x) !is.na(x) & nzchar(x)

raw    <- steps$prev_name != steps$src_name
folded <- raw & steps$prev_nm != steps$nm
named  <- folded & filled(steps$prev_nm) & filled(steps$nm)

funnel <- function(step, keep) {
  tibble::tibble(step = step, segments = sum(keep),
                 km = round(sum(steps$len_m[keep]) / 1000, 1))
}

bind_rows(
  funnel("written differently", raw),
  funnel("... and differ after folding", folded),
  funnel("... and neither side is blank", named))
```

Most of the raw churn is case, accents and numbered streets, which the files
write differently: 1996 writes `MARINE` and `15`, 2001 writes `Marine` and
`15th`. The matcher already folds all of this away. The blank-sided rows are roads that
gained or lost a name, which is a different event, so we remove them too.

```{r changes}
changes <- steps |> filter(named) |> rename(from = prev_nm, to = nm)
```

## The street type

The largest single pattern left is a street type appearing or disappearing at
the end of a name. The older files often write the type into the name field as
well as into `type`, and later ones stop. We can check this directly.

```{r type-moved}
with_type <- function(year, nm) {
  sources |>
    filter(vintage == year) |>
    inner_join(get_road_network(year, quiet = TRUE) |>
                 select(source_file, source_id, type),
               by = c("source_file", "source_id")) |>
    filter(src_name == nm) |>
    count(vintage, src_name, type) |>
    arrange(desc(n)) |>
    collect()
}

bind_rows(with_type(1991, "GRANDVIEW HIGHWAY"), with_type(1996, "GRANDVIEW"),
          with_type(1986, "EAST BOULEVARD"),    with_type(1991, "EAST"))
```

1991 writes Grandview Highway's type twice, once spelled out in `name` and once
as `HY` in `type`, whereas 1996 keeps only the second. East Boulevard in Kerrisdale is
the same move one census earlier, `BOULEVARD` in the name giving way to `BV` in
`type`. Same road, different filing.

## Peeling off the conventions

Three rules take out the rest of the changes that are about spelling rather than
naming: a deeper normalization that also ignores spacing, punctuation,
spelled-out numbers and `NO.` prefixes (`SOUTH RIDGE`/`SOUTHRIDGE`,
`133 A`/`133A`, `TENTH`/`10th`, `NO. 5`/`5`) and tolerates the Street Network
File's fixed-width name field, which truncates at twenty characters
(`COLUMBIA QUEBEC CONN`/`COLUMBIA QUEBEC CON.`); the trailing street-type word
above; and a distance cut, because a match 30 m away under a 38 m tolerance is
about as likely to be the next street over as the same one.

```{r classify}
spelled <- c(ZERO = 0, FIRST = 1, SECOND = 2, THIRD = 3, FOURTH = 4, FIFTH = 5,
             SIXTH = 6, SEVENTH = 7, EIGHTH = 8, NINTH = 9, TENTH = 10,
             ELEVENTH = 11, TWELFTH = 12)

deep <- function(x) {
  y <- gsub("\\bNO\\.? ", "", x)
  for (w in names(spelled)) y <- gsub(paste0("\\b", w, "\\b"), spelled[[w]], y)
  y <- gsub("\\b([0-9]+)(ST|ND|RD|TH)\\b", "\\1", y)
  gsub("[^A-Z0-9]", "", y)
}

# One name is the other cut short: the SNF name field is twenty characters wide.
clipped <- function(a, b) {
  k <- pmin(nchar(a), nchar(b))
  k >= 15 & substr(a, 1, k) == substr(b, 1, k)
}

types <- c("HIGHWAY", "HWY", "FREEWAY", "FRWY", "ROAD", "RD", "STREET", "ST",
           "AVENUE", "AVE", "DRIVE", "DR", "WAY", "BOULEVARD", "BLVD",
           "PARKWAY", "PKWY", "PKY", "CRESCENT", "CRES", "PLACE", "PL", "LANE",
           "LN", "COURT", "CRT", "CT", "TRAIL", "BRIDGE", "CAUSEWAY", "VIADUCT",
           "DIVERSION", "BYPASS", "BY-PASS", "MALL", "WALK", "WYND", "SQUARE",
           "ALLEY", "CONNECTOR", "THRUWAY", "EXPRESSWAY", "ROUTE", "DRIVEWAY")

# The one word a name has that the other does not, when that is the only change.
odd_word <- function(a, b) {
  mapply(function(x, y)
           if (length(x) == length(y) + 1L && identical(x[seq_along(y)], y))
             x[length(x)] else NA_character_,
         strsplit(a, " ", fixed = TRUE), strsplit(b, " ", fixed = TRUE),
         USE.NAMES = FALSE)
}

changes <- changes |>
  mutate(word = coalesce(odd_word(from, to), odd_word(to, from)),
         kind = case_when(
           deep(from) == deep(to)          ~ "spelling convention",
           clipped(deep(from), deep(to))   ~ "spelling convention",
           word %in% types                 ~ "street type moved",
           prev_dist >= 10 | dist_m >= 10  ~ "loosely matched",
           TRUE                            ~ "candidate rename"))

changes |>
  group_by(kind) |>
  summarize(segments = n(), km = round(sum(len_m) / 1000, 1)) |>
  arrange(desc(segments))
```

`loosely matched` is much the largest class here, and it is not a failure of the
files so much as of the question: downtown Vancouver puts named streets within
one tolerance of each other, so a cross-vintage match at 20 m is not evidence
that the two arcs are the same road.

## Names that change back

A name that changes and then changes back is a disagreement between files, not
an event on the ground:

```{r reversals}
pairs <- changes |>
  filter(kind == "candidate rename") |>
  group_by(from, to) |>
  summarize(segments = n(), km = round(sum(len_m) / 1000, 2), .groups = "drop")

reversed <- pairs |>
  semi_join(pairs |> select(from = to, to = from), by = c("from", "to"))

nrow(reversed)

reversed |> arrange(desc(km)) |> head(8)
```

`KENT` against `NORTH KENT` and `SOUTH KENT` is 1996 alone using a distinction
its neighbours do not, and `EAST BOULEVARD` against `MAPPLE` is 1976 alone, an
error the files on either side of it do not share. `ANDERSON` against
`GRANVILLE` is the files disagreeing about where one named structure ends and
the next begins.

## What is left

```{r stable}
renames <- pairs |>
  anti_join(reversed, by = c("from", "to")) |>
  arrange(desc(km))

tibble::tibble(pairs = nrow(renames), km = round(sum(renames$km), 1))

renames |> head(20)
```

Forty-two kilometres over fifty years, and it reads as a plausible list of one
city's renamings. `KENT` → `E KENT` and `W KENT` is Kent Avenue splitting;
`CONNAUGHT BRIDGE` → `CAMBIE BRIDGE` is the Cambie Street Bridge losing the name
it was opened under; `PARK DRIVEWAY` → `STANLEY PK DRIVEWAY` → `STANLEY PARK`
and `LOST LAGOON` → `N LAGOON` are Stanley Park's internal roads being
renamed twice; `OAK BRIDGE` → `OAK STREET BRIDGE` → `99` and `GRANVILLE` → `99`
are the highway number displacing the street name.

There is still room for improvement: `LANEYS MILL` → `LAMEY'S MILL` is a
Statistics Canada spelling error being fixed, and further cleaning rules could
remove cases like it.

## Dating a renaming

The transition columns say *between* which two vintages a name changed, which is
as fine as this data gets:

```{r kent-when}
kent <- changes |>
  filter(kind == "candidate rename", grepl("KENT", from) | grepl("KENT", to))

kent |> count(prev_year, vintage, from, to, sort = TRUE) |> head(8)
```

The 1996 excursion is one file's; the split into East and West Kent Avenue is a
single event in the 2011--2016 window, 110 segments of it at once. Mapped
against every stable renaming in the city, coloured by the file in which the new
name first appears:

```{r rename-map, fig.height = 5, fig.alt = "Map of the City of Vancouver street network in grey with renamed segments picked out in colour, keyed by the vintage in which the new name first appears; the longest coloured run is Kent Avenue along the Fraser River, renamed in 2016"}
stable <- changes |>
  filter(kind == "candidate rename") |>
  anti_join(reversed, by = c("from", "to"))

geo <- get_temporal_network("vancouver", within = van_city) |>
  filter(segment_id %in% !!stable$segment_id) |>
  collect_road_network() |>
  inner_join(stable |> select(segment_id, vintage), by = "segment_id") |>
  sf::st_transform(vancouver_crs)

context <- get_temporal_network("vancouver", within = van_city) |>
  filter(list_contains(years, 2021L)) |>
  collect_road_network() |>
  sf::st_transform(vancouver_crs)

ggplot(geo) +
  geom_sf(data = context, colour = "grey86", linewidth = 0.12) +
  geom_sf(aes(colour = factor(vintage)), linewidth = 0.8) +
  scale_colour_viridis_d(option = "turbo", direction = -1,
                         begin = 0.08, end = 0.94) +
  coord_sf(datum = NA) +
  labs(title = "Where the City of Vancouver's street names changed",
       subtitle = "Renamed segments, coloured by the first file to carry the new name",
       colour = "New name\nfirst appears",
       caption = "Source: Statistics Canada Area Master, Street Network and Road Network Files") +
  guides(colour = guide_legend(override.aes = list(linewidth = 1.2)))
```

Renaming is thin, and it clusters: the waterfronts, the bridges and viaducts,
Stanley Park, and the long 2016 run of Kent Avenue along the Fraser.

## Caveats

The vignette demonstrates a basic workflow; there are several limits to keep in
mind.

**A road that was renamed *and* redrawn is invisible.** The geometry match needs
the two arcs within the calibrated tolerance; the name rescue needs the names to
agree exactly.
A road that fails both is emitted as one segment retiring and another
appearing, with nothing connecting them. That is the systematic blind spot, and
it falls hardest on the long arcs the older files generalize into a coarse
chord --- a smaller problem inside a dense city grid than it would be across the
region.

**The resolution is the vintage spacing.** Every date here is a five-year
interval, the gap between one census and the next.

**A change in the record is not a change on the ground.** `LANEYS MILL` →
`LAMEY'S MILL` is a correction, and the whole `loosely matched` class above is
neither a correction nor a renaming. Separating a municipal renaming from either
needs a source this data does not contain.

## Attribution

Source data are © Statistics Canada, distributed under the
[Statistics Canada Open Licence](https://www.statcan.gc.ca/en/reference/licence).
Cite the product and reference year in anything you publish from it.
