---
title: "Tracking the road network through time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tracking the road network through time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 5, dpi = 96, out.width = "100%",
                      fig.path = "canstreet-temporal-")
```

`get_road_network()` gives each census year on one schema, but the years do not
line up: the same road is cut into different arcs from one release to the next,
identifiers change scheme three times, and every arc is redrawn as digitizing
improves. `build_temporal_network()` reconciles them into one table in which
each segment carries the years it is present in.

This vignette is precomputed, but the code is what you would run.

```{r library}
library(canstreet)
library(dplyr)
library(ggplot2)
```

## A region to work in

`within` takes any `sf`, `sfc` or `bbox` object, and the same polygon is stamped
across every year: Statistics Canada reuses `CSDUID` and `CMAUID` across
boundary revisions, and the 1996 and 2006 files carry no region columns at all.
Any source of a polygon will do --- here is the 2021 Calgary CMA from
`cancensus`, which needs no API key for geographies:

```{r region}
cma <- cancensus::get_statcan_geographies("2021", level = "CMA") |>
  filter(CMANAME == "Calgary")

cma |> select(CMAUID, CMANAME) |> sf::st_drop_geometry()
```

## Building

Name the years --- any two or more --- and anything not yet imported is
downloaded first. `NULL` uses whatever the cache holds.

```{r build, message = FALSE}
calgary <- build_temporal_network(
  "calgary", c(1996, 2001, 2006, 2011, 2016, 2021),
  within = cma, region_note = "2021 CMA boundary, Calgary (825)")

calgary
```

The result is a lazy table like any other in this package, so it collects with
`collect_road_network()` and exports with `export_road_network()`.

## How much road was there, when

`year_key` is the year set as a string, so "how many kilometres were present in
exactly these years" is a plain `GROUP BY`:

```{r year-key}
calgary |>
  group_by(year_key) |>
  summarize(km = round(sum(len_m, na.rm = TRUE) / 1000, 1)) |>
  arrange(desc(km)) |>
  collect()
```

`years` is the same set as a list column, for how much road existed in a given
year:

```{r by-year}
vints <- c(1996, 2001, 2006, 2011, 2016, 2021)

tibble::tibble(year = vints, km = round(vapply(vints, function(y) {
  calgary |>
    filter(list_contains(years, y)) |>
    summarize(km = sum(len_m, na.rm = TRUE) / 1000) |>
    pull(km)
}, numeric(1)), 1))
```

1996 is urban areas only and 2001 is the first national file, so the jump
between them is mostly coverage, not road being built.

`first_year`, `last_year` and `n_years` fall out of the year set:

```{r new-and-gone}
calgary |>
  summarize(
    new_since_2006_km = sum(ifelse(first_year > 2006, len_m, 0),
                            na.rm = TRUE) / 1000,
    retired_km        = sum(ifelse(last_year < 2021, len_m, 0),
                            na.rm = TRUE) / 1000) |>
  collect()
```

## Newer geometry wins

A segment's geometry is cut from the newest year it appears in: the newest
vintage is laid down as the spine, and each earlier year contributes only where
no later one covers it. So `spine_vintage` always equals `last_year`:

```{r spine}
calgary |>
  count(spine_vintage, last_year) |>
  arrange(spine_vintage) |>
  collect()
```

## Segments are cut where the years disagree

The vintages disagree on where one arc ends and the next begins, so each spine
arc is cut wherever another year's coverage changes. The build therefore holds
more segments than any single year holds arcs:

```{r arcs}
get_road_network(vints, within = cma) |>
  count(vintage) |>
  arrange(vintage) |>
  collect()

calgary |> summarize(harmonized_segments = n()) |> collect()
```

## How the tolerance was chosen

The tolerance is measured, not picked: how far apart two files put the same
named road, against the rate at which a tolerance starts matching the
neighbouring road instead.

```{r calibration}
temporal_network_calibration("calgary")
```

`recall_p50` is the positional disagreement against 2021; `disagree_25` the
share of arcs whose nearest arc within 25 m carries a different name. The
tolerance is set where the second is still flat, and `tolerance =` overrides it
as a number or a vector named by year. 2016 clamps to the lower bound because
it shares 2021's base geometry.

## The matching, exposed

`get_temporal_network_sources()` is the crosswalk --- one row per segment per
year, naming the arc of that year's file it corresponds to, how far away it
was, and which rule matched it:

```{r crosswalk}
get_temporal_network_sources("calgary") |>
  count(vintage, match_kind) |>
  arrange(vintage, desc(n)) |>
  collect()
```

`spine` is the year the geometry came from; `geometry` and `geometry+name` are
close approaches that also agree on direction. `name_rescue` is the second
rule, for the long rural and highway arcs an older file draws as a chord
hundreds of metres off the modern curve --- without it Calgary reports some
1,180 km of spurious retirement. It is recorded rather than hidden, so a
stricter analysis can drop it:

```{r rescue-share}
get_temporal_network_sources("calgary") |>
  filter(vintage == 2006) |>
  summarize(rescued = sum(ifelse(match_kind == 'name_rescue', 1, 0),
                          na.rm = TRUE),
            total = n()) |>
  collect()
```
The harmonized network, coloured by the year each segment first appears:

```{r first-year-map, fig.alt = "Calgary CMA road network coloured by the first vintage each segment appears in"}
colours <- c("grey40", viridis::viridis(length(vints) - 1, option = "turbo",
                                        direction = -1, begin = 0.12, end = 0.92))

get_temporal_network("calgary") |>
  filter(list_contains(years, 2021L)) |>
  collect_road_network() %>%
  sf::st_transform(mountainmathHelpers::lambert_conformal_conic_at(.)) |>
  ggplot(aes(colour=factor(first_year))) +
  geom_sf(linewidth=0.1) +
  scale_colour_manual(values = colours) +
  labs(title = "Calgary CMA road network by vintage",
       colour="First year\nroad appears\nin data",
       caption="StatCan Road/Street network files") +
  guides(colour=guide_legend(override.aes=list(linewidth=1))) +
  coord_sf(datum=NA) 
```



## What actually disappeared

Road that was there in 2006 and is not there now:

```{r retired}
retired <- get_temporal_network("calgary") |>
  filter(list_contains(years, 2006L), last_year < 2021) |>
  collect_road_network()

sum(retired$len_m) / 1000
```

```{r retired-names}
retired |>
  sf::st_drop_geometry() |>
  mutate(name = ifelse(is.na(name) | name == "", "(unnamed)", name)) |>
  group_by(name) |>
  summarize(km = round(sum(len_m) / 1000, 2)) |>
  arrange(desc(km)) |>
  head(10)
```

```{r map, fig.alt = "Roads present in 2006 and gone by 2021, over the Calgary CMA"}
get_temporal_network("calgary") |>
  filter(list_contains(years, 2021L)) |>
  collect_road_network() %>%
  sf::st_transform(mountainmathHelpers::lambert_conformal_conic_at(.)) |>
  ggplot() +
  geom_sf(fill = "grey85", color = "black", linewidth=0.1) +
  geom_sf(data = retired, color = "firebrick", linewidth = 0.3) +
  labs(title = "Calgary CMA: road present in 2006, gone by 2021") +
  coord_sf(datum=NA)

```

Retirement is a weaker claim than growth: a road the matcher missed looks
identical to one that was removed. Segments whose year set has an interior gap
--- present in 1996 and 2021 but not 2006 --- measure how often that happens,
since a road cannot un-exist and come back:

```{r gaps}
by_key <- calgary |>
  group_by(year_key, first_year, last_year, n_years) |>
  summarize(km = sum(len_m, na.rm = TRUE) / 1000, .groups = "drop") |>
  collect()

# How many of the build's years lie between this segment's first and last?
expected <- colSums(outer(vints, by_key$first_year, ">=") &
                    outer(vints, by_key$last_year, "<="))
gap_km <- sum(by_key$km[by_key$n_years < expected])

c(gap_km = round(gap_km, 1), share = round(gap_km / sum(by_key$km), 4))
```

## Regions move under the roads

The crosswalk records each year's own `csduid_l` against a segment whose
geometry is fixed, so every row here is an annexation, a boundary revision or a
recoding:

```{r drift}
temporal_network_region_drift("calgary") |>
  mutate(km = round(km, 1)) |>
  head(8)
```

Only years carrying the attribute contribute; 1996, 2001 and 2006 have none.

## Builds in the cache

```{r listing}
list_temporal_networks() |>
  select(name, vintages, n_segments, total_km, regional, built_at)
```

A build is a pair of tables beside the vintages, so `remove_temporal_network()`
drops it instantly, and removing a vintage removes the builds that used it.

`within = NULL` builds nationally, but that is not yet the supported scale: a
national two-vintage build spilled over 32 GB of DuckDB temporary storage here
and did not finish. Use region-scoped builds.

## Attribution

Source data are © Statistics Canada, distributed under the
[Statistics Canada Open Licence](https://www.statcan.gc.ca/en/reference/licence).
Cite the product and reference year in anything you publish from it.
