---
title: "Tracking the road network through time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tracking the road network through time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 5, dpi = 96, out.width = "100%",
                      fig.path = "canstreet-temporal-")
```

`get_road_network()` gives each census year on one schema, but the years do not
line up: the same road is cut into different arcs from one release to the next,
identifiers change scheme three times, and every arc is redrawn as digitizing
improves. `build_temporal_network()` reconciles them into one table in which
each segment carries the years it is present in.


```{r library, message = FALSE}
library(canstreet)
library(dplyr)
library(ggplot2)
```

## Calgary as example

Operating on the entire country at once is very resource intensive, so temporal
harmonization is best done one region at a time.

`within` takes any `sf`, `sfc` or `bbox` object, and the same polygon is stamped
across every year: Statistics Canada reuses `CSDUID` and `CMAUID` across
boundary revisions, and the files before 2011 carry no region columns at all.
For this we use the 2021 Calgary CMA from
`cancensus`, which needs no API key for geographies:

```{r region}
cma <- cancensus::get_statcan_geographies("2021", level = "CMA") |>
  filter(CMANAME == "Calgary")

cma |> select(CMAUID, CMANAME) |> sf::st_drop_geometry()
```

## Building

We use all census-year vintages. Passing `NULL` instead uses whatever the cache
holds; a vintage named explicitly that is not in the cache yet is imported
automatically in the process.

```{r build, message = FALSE}
vints <- c(1971, 1976, 1981, 1986, 1991, 1996, 2001, 2006, 2011, 2016, 2021)

calgary <- build_temporal_network(
  "calgary", vints, within = cma,
  region_note = "2021 CMA boundary, Calgary (825)")

calgary
```

Building over the Area Master Files warns once for each of their four years,
1971 to 1986, that same-named arcs are systematically displaced; the calibration
section below comes back to it. A build already in the cache is returned as it
stands, without rebuilding or warning again, and `refresh = TRUE` forces one.

The result is a lazy table like any other in this package, so it collects with
`collect_road_network()` and exports with `export_road_network()`.

## How much road was there, when

`year_key` is the year set as a string, so "how many kilometres were present in
exactly these years" is a plain `GROUP BY`:

```{r year-key}
calgary |>
  group_by(year_key) |>
  summarize(km = round(sum(len_m, na.rm = TRUE) / 1000, 1)) |>
  arrange(desc(km)) |>
  collect()
```

`years` is the same set as a list column, for how much road existed in a given
year:

```{r by-year}
tibble::tibble(year = vints, km = round(vapply(vints, function(y) {
  calgary |>
    filter(list_contains(years, y)) |>
    summarize(km = sum(len_m, na.rm = TRUE) / 1000) |>
    pull(km)
}, numeric(1)), 1))
```

Everything up to 1996 is urban areas only and 2001 is the first national file,
so the jump between them is mostly coverage, not road being built. The steadier
climb before it is a statement about the files as well: each holds the urban
area as that census drew it.

`first_year`, `last_year` and `n_years` fall out of the year set:

```{r new-and-gone}
calgary |>
  summarize(
    new_since_2006_km = sum(ifelse(first_year > 2006, len_m, 0),
                            na.rm = TRUE) / 1000,
    retired_km        = sum(ifelse(last_year < 2021, len_m, 0),
                            na.rm = TRUE) / 1000) |>
  collect()
```

## Newer geometry wins

A segment's geometry is cut from the newest year it appears in: the newest
vintage is laid down as the spine, and each earlier year contributes only where
no later one covers it. So `spine_vintage` always equals `last_year`:

```{r spine}
calgary |>
  count(spine_vintage, last_year) |>
  arrange(spine_vintage) |>
  collect()
```

## Segments are cut where the years disagree

The vintages disagree on where one arc ends and the next begins, so each spine
arc is cut wherever another year's coverage changes. The build therefore holds
more segments than any single year holds arcs:

```{r arcs}
get_road_network(vints, within = cma) |>
  count(vintage) |>
  arrange(vintage) |>
  collect()

calgary |> summarize(harmonized_segments = n()) |> collect()
```

## How the tolerance was chosen

The tolerance is measured, not picked: how far apart two files put the same
named road, against the rate at which a tolerance starts matching the
neighbouring road instead.

```{r calibration}
temporal_network_calibration("calgary") |>
  select(vintage, tolerance_m, n_pairs, recall_p50, disagree_25, dx, dy)
```

`recall_p50` is the positional disagreement against 2021; `disagree_25` the
share of arcs whose nearest arc within 25 m carries a different name. The
tolerance is set where the second is still flat, and `tolerance =` overrides it
as a number or a vector named by year. 2016 clamps to the lower bound because
it shares 2021's base geometry.

`dx` and `dy` are the mean offset to the same-named 2021 arc, in metres along
the axes of the storage projection, and they are what the build warned about.
A few metres either way is ordinary. The four Area Master File years sit a
steady 11 to 15 m off, all in the same direction, which is a difference in
registration rather than noise. The matcher does not correct a shift, it absorbs
it --- those years calibrate to the 40 m ceiling, with `disagree_25` well above
the later files' --- so treat a difference at that scale in 1971 to 1986 as
unresolved. The shift present in Calgary is not universal: Vancouver's Area Master
Files register as well as any file before 2011 (`vignette("canstreet-vancouver")`).

## Road matching details

`get_temporal_network_sources()` is the crosswalk --- one row per segment per
year, naming the arc of that year's file it corresponds to, how far away it
was, and which rule matched it:

```{r crosswalk}
get_temporal_network_sources("calgary") |>
  count(vintage, match_kind) |>
  arrange(vintage, desc(n)) |>
  collect()
```

`spine` is the year the geometry came from; `geometry` and `geometry+name` are
close approaches that also agree on direction. `name_rescue` is the second
rule, for the long rural and highway arcs an older file draws as a chord
hundreds of metres off the modern curve --- without it Calgary reports some
1,180 km of its 2006 road as retired. It is recorded rather than hidden, so a
stricter analysis can drop it:

```{r rescue-share}
get_temporal_network_sources("calgary") |>
  filter(vintage == 2006) |>
  summarize(rescued = sum(ifelse(match_kind == 'name_rescue', 1, 0),
                          na.rm = TRUE),
            total = n()) |>
  collect()
```

## Mapping the results

The harmonized network, coloured by the year each segment first appears. Most of
the CMA's area is rural grid, so the maps here are framed on the City of Calgary.

```{r first-year-map, fig.height = 7.5, fig.alt = "City of Calgary road network coloured by the first vintage each segment appears in, 1971 to 2021"}
calgary_crs <- mountainmathHelpers::lambert_conformal_conic_at(cma)

frame <- cancensus::get_statcan_geographies("2021", level = "CSD") |>
  filter(CSDUID == "4806016") |>
  sf::st_transform(calgary_crs) |>
  sf::st_bbox()

colours <- c("grey40", viridis::viridis(length(vints) - 1, option = "turbo",
                                        direction = -1, begin = 0.12, end = 0.92))

get_temporal_network("calgary") |>
  filter(list_contains(years, 2021L)) |>
  collect_road_network() |>
  sf::st_transform(calgary_crs) |>
  ggplot(aes(colour = factor(first_year))) +
  geom_sf(linewidth = 0.15) +
  scale_colour_manual(values = colours) +
  labs(title = "Calgary road network by vintage",
       colour = "First year\nroad appears\nin data",
       caption = "StatCan Area Master, Street Network and Road Network Files") +
  guides(colour = guide_legend(override.aes = list(linewidth = 1))) +
  coord_sf(datum = NA, xlim = frame[c("xmin", "xmax")],
           ylim = frame[c("ymin", "ymax")])
```

The grey core marks the 1971 start of the data series and the rings around it are the urban files following the city outward to 1996; the green grid at the edges of the
frame is 2001 going national.

## Roads that disappeared

Restricting ourselves to the national data from 2006 onward (or alternatively to
a sub-region that all vintages cover) lets us look at roads that are present in
2006 but not in the 2021 data.

```{r retired}
retired <- get_temporal_network("calgary") |>
  filter(list_contains(years, 2006L), last_year < 2021) |>
  collect_road_network()

sum(retired$len_m) / 1000
```

```{r retired-names}
retired |>
  sf::st_drop_geometry() |>
  mutate(name = ifelse(is.na(name) | name == "", "(unnamed)", name)) |>
  group_by(name) |>
  summarize(km = round(sum(len_m) / 1000, 2)) |>
  arrange(desc(km)) |>
  head(10)
```

Most of those 713 km lie outside the city, on the rural grid --- about 150 km
are inside the city boundary. 

```{r map, fig.height = 8.5, fig.alt = "Roads present in 2006 and gone by 2021, over the City of Calgary"}
get_temporal_network("calgary") |>
  filter(list_contains(years, 2021L)) |>
  collect_road_network() |>
  sf::st_transform(calgary_crs) |>
  ggplot() +
  geom_sf(colour = "grey60", linewidth = 0.1) +
  geom_sf(data = retired, colour = "firebrick", linewidth = 0.4) +
  labs(title = "Calgary: road present in 2006, gone by 2021") +
  coord_sf(datum = NA, xlim = frame[c("xmin", "xmax")],
           ylim = frame[c("ymin", "ymax")])
```

One caveat is that a road may be misidentified as retired if the matcher missed
it, so these results need validation before anything relies on them.

A related check is the length with gaps: segments missing from a year that lies
between their first and last appearance.


```{r gaps}
by_key <- calgary |>
  group_by(year_key, first_year, last_year, n_years) |>
  summarize(km = sum(len_m, na.rm = TRUE) / 1000, .groups = "drop") |>
  collect()

# How many of the build's years lie between this segment's first and last?
expected <- colSums(outer(vints, by_key$first_year, ">=") &
                    outer(vints, by_key$last_year, "<="))
gap_km <- sum(by_key$km[by_key$n_years < expected])

c(gap_km = round(gap_km, 1), share = round(gap_km / sum(by_key$km), 4))
```

## Regions move under the roads

The crosswalk records each year's own `csduid_l` against a segment whose
geometry is fixed, so every row here is an annexation, a boundary revision or a
recoding:

```{r drift}
temporal_network_region_drift("calgary") |>
  mutate(km = round(km, 1)) |>
  head(8)
```

Only years carrying the attribute contribute, and nothing before 2011 does.

## Builds in the cache

```{r listing}
list_temporal_networks() |>
  select(name, vintages, n_segments, total_km, regional, built_at)
```

A build is a pair of tables beside the vintages, so `remove_temporal_network()`
drops it instantly, and removing a vintage removes the builds that used it.


## Attribution

Source data are © Statistics Canada, distributed under the
[Statistics Canada Open Licence](https://www.statcan.gc.ca/en/reference/licence).
Cite the product and reference year in anything you publish from it.
