---
title: "Getting started with canstreet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with canstreet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Figures are written beside the .Rmd rather than into a `figure/` directory:
# R CMD check reports a directory by that name in vignettes/ as knitr detritus,
# but a precomputed vignette has to ship its figures.
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96, out.width = "100%",
                      fig.path = "canstreet-")
```

This vignette is precomputed --- it downloads several hundred megabytes --- but
the code is what you would run.

```{r library, message = FALSE}
library(canstreet)
library(dplyr)
```

## Set a cache path

A national vintage is 200–350 MB compressed and about 2.2 million segments, so
`canstreet` caches the archives and the harmonized database. This cache path should be set as CANSTREET_CACHE_PATH environment, the `set_canstreet_cache_path` facilitates this. If the `install = TRUE` parameter is set it installes it into .Renviron so it will be available in all future sessions and does not need to get called again.

```{r cache, eval = FALSE}
set_canstreet_cache_path("~/data/canstreet", install = TRUE)
```


## Available vintages

```{r vintages}
list_road_network_vintages() |>
  select(vintage, product, coverage, source_crs, imported)
```

The 1971 to 1986 Area Master Files and the 1991 and 1996 Street Network Files
cover urban areas only; everything from 2001 on is national. Starting with 2005 data is available at yearly frequency, prior to that for census years only.

## Getting a vintage

The `get_road_network` is the universal way to access the road network data for a given vintage, if needed it fetches the data and parses it into a local database, and hands back a lazy DuckDB table.

```{r get}
roads <- get_road_network(2021)
roads
```

Once downloaded connecting to the database is instant, `dplyr` data operations should be chained before calling collect so they are run on DuckDB to keep memory overhead low and for optimized fast execution.

```{r by-province}
roads |>
  filter(!is.na(pruid_l)) |>
  group_by(pruid_l) |>
  summarize(segments = n(), km = round(sum(len_m, na.rm = TRUE) / 1000)) |>
  arrange(desc(km)) |>
  collect()
```

## Spatial filters

Pass any `sf`, `sfc` or `bbox` object as `within`. Over a single vintage the
filter uses that vintage's R-tree index, so it stays fast on a national file.
`collect_road_network()` then materializes the result as `sf`.

```{r vancouver}
bbox <- sf::st_bbox(c(xmin = -123.15, ymin = 49.26, xmax = -123.09,
                      ymax = 49.29), crs = 4326)

downtown <- get_road_network(2021, within = bbox) |>
  collect_road_network()

downtown |> select(name, type, af_l, at_l, csdname_l)
```

```{r plot, fig.alt = "Street segments in downtown Vancouver, 2021"}
plot(sf::st_geometry(downtown), col = "grey30", lwd = 0.6)
```

Geometry is always converted to and stored in EPSG:3347 (NAD83 / Statistics Canada Lambert), so `len_m`
is metres for every vintage. `collect_road_network()` reprojects on request:

```{r crs}
get_road_network(2021, within = bbox) |>
  collect_road_network(crs = 4326) |>
  sf::st_crs() |>
  (\(x) x$epsg)()
```

## Comparing vintages

Calling `get_road_network` with several vintages hands back a database connection with the individual vintages stacked that operates on all vintages simultaneously.

```{r across-years}
get_road_network(c(1971, 1976, 1981, 1986, 1991, 1996, 2001, 2006, 2011, 2016,
                   2021)) |>
  group_by(vintage) |>
  summarize(segments = n(), km = round(sum(len_m, na.rm = TRUE) / 1000)) |>
  arrange(vintage) |>
  collect()
```

Nothing before 2001 is national, and the Area Master Files grow mostly by
covering more places, 34 municipal files in 1971, 194 in 1986, so most of the rise
to 1996 is due to expanding coverage. Segment counts do not compare across
products either. 1986 holds more segments than 1991 over fewer kilometres
because an Area Master File segment is a single block face. And not every
segment is a road, over a third of 1986's arcs are shorelines, boundaries and
railways, and 2001's extra 400,000 km is census boundary topology.
Setting the `roads_only = TRUE` parameter drops both. The 2006–2021 drop is in the source data, not
the import. 2006 carries far more very long arcs, which later releases split,
reclassified or dropped.

```{r long-arcs}
get_road_network(c(2006, 2021)) |>
  filter(len_m > 5000) |>
  group_by(vintage) |>
  summarize(long_arcs = n(), km = round(sum(len_m, na.rm = TRUE) / 1000)) |>
  collect()
```

Identifiers are not stable across vintages, so comparing years
is done by matching on geometry only. `build_temporal_network()` does that and writes one
table in which each segment carries the years it is present in. For more details refer to vignette explaining the temporal matching 
`vignette("canstreet-temporal")`. 

## The harmonized schema

Every vintage maps onto the same columns, if a vintage does not carry a specific column the values are set to `NA`.

```{r schema}
canstreet_schema()
```

Source spellings differ by vintage — `arc_id` in 1991/1996, `RB_UID` in
2001–2010, `NGD_UID` from 2011, `NGDUID` in 2016/2017 — and the importer
resolves each by alias.

## Exporting

For use in other systems like Python or QGIS one can use the same DuckDB, the local path can be accessed via the `path` property of `get_road_network_database()`. There also is a `export_road_network()` convenience function to explicitly export the data to GeoParquet with the CRS in its metadata:

```{r export, eval = FALSE}
get_road_network(2021, within = bbox) |>
  export_road_network("vancouver_2021.parquet")
```

## Managing the cache

Several convenience functions help manage the cache.

```{r cache-listing}
list_canstreet_cache()
```

`remove_canstreet_cache(vintage)` drops a vintage's table and keeps its
archives, so it can be re-imported without downloading again; pass
`keep_raw = FALSE` to delete those too.

## Attribution

Source data are © Statistics Canada, distributed under the
[Statistics Canada Open Licence](https://www.statcan.gc.ca/en/reference/licence).
Cite the product and reference year in anything you publish from it.
